---
title: "BIOS 512- Homework 2"
author: "Alexis Bryant"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
  word_document: default
---

```{r, results="hide", message=FALSE, warning=FALSE}
library(tidyverse);
 tinytex::install_tinytex(force = TRUE)
```
## Question 1 

What are three ways that we can assign the value of 25 to y?

```{r}
y <- 25
y=25
y <<- 25

y
```
## Question 2

How would I print the following line, including the quotes, in R?\
**He said, "I'm Garth Marenghi. Author. Dreamweaver. Visionary. Plus actor."**

```{r}
cat("He said, \"I'm Garth Marenghi. Author. Dreamweaver. Visionary. Plus actor.\"")
```
## Question 3 

What if I wanted to add backslashes into the statement from Question 1 to make the statement below?\
**He said, "I'm Garth Marenghi. Author\\Dreamweaver\\Visionary. Plus actor."**

```{r}
cat("He said, \"I’m Garth Marenghi. Author\\Dreamweaver\\Visionary. Plus actor.\"")
```
## Question 4

Show two ways to get the following array: 1 2 3

```{r}
1:3
c(1,2,3)
```
## Question 5

What does R call things like +, -, sin(), c(), etc? What about \<-?

+,-,sin(),c() are functions and <- is an assignment operator or function/special operator that creates variable bindings

## Question 6

What's the difference in the way R processes the while() and the for() below?

```{r}
x <- 0;
while (x <= 3) {
    x <- x + 1;
}
x

for(y in c(1,2,3)) {
     y <- y + 1;
}
y


```

In the while loop R reevaluates the function condition each time and updates x in place until the condition fails.
The for loop iterates over the values given in the sequence and inside each iteration, y is binded to the current element and then it changes the loop variable. After the for loop, y just keeps the last changed value.
So, while loops depend on a loop condition and continuously updates the variable binding and for loops iterate across a sequence of values where it binds the loop variable each time.

## Question 7

Create the Pythagorean formula and evaluate it with a=3 and b=4. Print the output.

```{r}
a <- 3
b <- 4
c <- sqrt(a^2 + b^2)
c
```
## Question 8

Load the help for the built in sin() function.

```{r}
?sin
```
## Question 9

Which version of the counter function works? What is the difference in the way R processes the two functions?

```{r}
counter1 <- function(start, step){
    val <- start;
    function(){
        old_val <- val;
        val <- val + step;
        old_val;
    }
}
counter_from_1 <- counter1(1,1);
counter_from_1()
counter_from_1()

counter2 <- function(start, step){
    val <- start;
    function(){
        old_val <- val;
        val <<- val + step;
        old_val;
    }
}
counter_from_1 <- counter2(1,1);
counter_from_1()
counter_from_1()

```


Counter 2 works because it uses <<- the super assignment so it modifies the closest binding in the original environment so the value gets updated
Counter 1 doesn't work because it uses <- in the inside function, so it creates a new variable binding instead of updating the other val variable so it keeps returning 1


## Question 10

a)  Use read_csv the cars.csv and store it in a data frame.
b)  Then, group the data frame by Make, get averages across the numeric variables, and then sort by Volume in descending order. Hint: Use summarise(across(c(<variable list>), mean)) to get the averages.

```{r (a)}
library(tidyverse)

df.cars <- read_csv("cars.csv")
```
```{r (b)}
cars_summary <- df.cars %>%
    group_by(Make) %>%
  summarise(
      across(c(Volume, Weight, CO2), mean))%>%
  arrange(desc(Volume))

cars_summary
```
## Question 11

Make a function that returns the Fibonacci sequence, then call it 7 times to return the first 7 values of the sequence. Use the correct counter function from question 9 for inspiration.

```{r}
fibonacci_Sequence <- function(a0 = 0, b0 = 1) {
  a <- a0; b <- b0
  function() {
    output <- a              
    next_b <- a + b        
    a <<- b              
    b <<- next_b
    output
  }
}

fib_seq <- fibonacci_Sequence(0, 1)
vals <- vapply(1:7, function(i) fib_seq(), numeric(1))
vals

```

